---
title: "Baseball yay"
subtitle: "Programming in Statistics - Project"
author: "Vilsmeier Johannes, Sterneder Florian, Schaub Andrea"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    theme: cosmo
    highlight: tango
    number_sections: true
    toc: true
    toc_float: yes
    fig_width: 7
    fig_height: 4.5
---

```{r hide=T, warning=F, message=F}

library(knitr)
opts_chunk$set(echo = TRUE,
               collapse = TRUE,
               comment = "",
               strip.white = TRUE,
               warning = FALSE,
               messages = FALSE,
               out.width = "70%",
               fig.align = "center")
source("help.R")
```  
#Proportions of Batting And Throwing Hands Since 1871  

##Data Preparation   

Our aim was to compute and plot the number of left, right and - if applicable - both-handed players that were in the league in any given year. The ```People``` table from Lahman's Baseball database contains information on each player's preferred batting and throwing hand, while the ```Appearances``` table contains information on which players were active each year, since 1871. Thus, if a player was active in, say 1871, he should appear in the ```Appearances``` table and given the information on his preferred batting and throwing hand from the ```People``` table, we should be able to derive the number of right, left, or both-handed players that were active in a given year. 

Unfortunately, a simple join of the ```People```and ```Appearances``` tables via the intended key "playerID" is not possible, since a given player can be listed multiple times in the ```Appearances``` table. This is usually due to players switching teams midseason. If a player switched teams, he will appear as many times in the ```Appearances``` table as the number of teams he played for that year. It follows that if we joined ```Appearances```and ```People``` and computed the number/proportion of players in each handedness category and year, we would count some players more than once. 

```{r}
#Check number of observations per player and year
Appearances %>% 
  count(playerID, yearID) %>% 
  filter(n > 1)
```  
We solved this issue by creating a new table ```didAppear``` which indicates whether a player appeared at least once in a given year (i.e., there's only one row per year and player).

```{r}
didAppear <- 
  Appearances %>% 
  group_by(yearID) %>% 
  distinct(playerID) %>% 
  ungroup()

didAppear
```

Now, each player is counted only once per active year. 
```{r}
didAppear %>% 
  count(playerID, yearID) %>% 
  filter(n > 1)
```  
The ```People``` table already has a unique key in "playerID".

```{r}
People %>% 
  count(playerID) %>% 
  filter(n > 1)
```


We used ```left_join()``` to combine the ```didAppear``` and the ```People``` tables. The variable "playerID" served as a key. In the resulting table ```players``` each row corresponds to a unique year-by-player combination. A players preferred batting hand is encoded in the variable "bats", while his preferred throwing hand is encoded in the variable "throws". We changed these variables from type "character" to type "factor" and ordered their possible outcomes as follows: Right < Left < Both for "bats", and Right < Left, for "throws" (Note that bats had 3 possible outcomes, while throws only had 2). This ordering ensured that the legends in the final plots were ordered from most frequent to least frequent. 

Unfortunately, the "throws" variable contained a single observation with the value "S" (instead of "R", or "L"). Since there was no mentioning of outcome "S" in the description of Lahman's baseball database, we concluded that this must be a typo or some other mistake. Therefore "S" was changed to "NA". 

```{r}
unique(People$throws) #possible outcomes for throws
sum(People$throws == "S", na.rm = T) #Count number of observations with "S"
People$throws[People$throws=="S"] <- NA

``` 


```{r}
players <- left_join(didAppear,People,by="playerID") %>% 
  mutate(
    bats = factor(bats, 
                  levels = c("R","L","B"),
                  labels= c("Right","Left","Both"), #3 possible outcomes
                  ordered=T),#To get the legend of the plot in the right order
    throws = factor(throws,
                    levels = c("R","L"),
                    labels = c("Right", "Left"), #only 2 possible outcomes
                    ordered = T)#To get the legend of the plot in the right order
  ) %>% 
  rename("year" = yearID)

#check if each playerID-by-year combination is unique
players %>% 
  count(playerID, year) %>% 
  filter(n > 1)
```

##Batting Hand
The table ```plrBats```contains each year's total number of observations for all three handedness categories (left, right, both; "observed"), the total number of observations, irrespective of handedness category ("N"), as well as the proportion of each handedness category ("proportion"). 
```{r}
plrBats <- players %>% 
  group_by(year,bats) %>% #Ensures that we get the data for each year
  count() %>%
  ungroup() %>% 
  drop_na() %>% #Drop observations for which the batting hand was not available
  group_by(year) %>% 
  mutate(
    "N" = sum(n), #Get total number of observations for each year (irrespective of handedness category) 
    "proportion" = round(n/sum(n),3)
  ) %>% 
  rename("observed" = n)
plrBats
```

##Plot  
We figured that the number of observations a given year's proportional estimate is based upon may be an important piece of information. In order to include this information, we used ```ggplotly()``` as it allowed us to include N in ```plotly```'s own "hover text". 
In more detail, we first created the ```ggplot```object ```batsPlot```  using ```geom_lines```to connect each year's proportions, with separate lines for each handedness category. Notice that we specified a third aesthetic "label = N" in addition to the x- and y-coordinates. This third aesthetic was necessary, as the ```tooltip```argument from ```ggplotly()``` can only display data that are specified as aesthetic arguments in the ```ggplot```object. 
If you hover your cursor over any data point in the plot, a text element should pop up, displaying the year, the proportion and the total number of observations, N. To specify which aesthetics are displayed by ```ggplotly```, we provided the  ```tooltip``` argument with a string vector of the follwoing aesthetics: "year", "proportion", "N". 


```{r}
batPlot <- 
  #aesthetic "label" is necessary for ggplotly's tooltip argument
  ggplot(plrBats, aes(year, proportion, label = N))+ 
  geom_line(aes(color=bats))+
  scale_x_continuous(name="Year", breaks = seq(1871,2018,20))+
  scale_y_continuous(breaks=seq(0,1,0.1))+
  scale_color_brewer(palette="Set1")+
  ylab("Proportion of Preferred Batting Hand")+
  labs(color="Batting \n Hand")+
  ggtitle("Proportional Distribution of Batting Hands")

#Convert ggplot object into plotly object
ggplotly(batPlot, tooltip = c("proportion", "year", "N")) 

``` 
The plot may have us think that the proportion of right handed batters first increased steeply between 1871 and 1876 (up to .88, while in the same time span, the proportion of left handed batters decreased in a similar fashion) and then decreased again until settling somewhere between 0.6 and 0.7 between 1900 and 2018. However, this early rise and fall in the proportion of right handed batters does not necessarily represent a genuine trend but may simply reflect the degree of uncertainty in the proportional estimates that come with relatively small sample sizes, especially in the early years. For instance, the proportional estimates for 1871 are based on a total of $N = 48$ observations and if we take a look at the actual number of observations within each handedness group we will notice that there was only **one** both handed batter in our data frame in 1871.

##Throwing Hand  

To obtain the proportions of left- and right-handed throwers for each year, we proceeded analogously to the more detailed account above. The table ```plrThrows``` contains the each year's total number of observations both handedness categories (left and right; "observed"), the total number of observations, irrespective of handedness category ("N"), as well as the proportion of each handedness category ("proportion"). 

```{r}
plrThrows <- players %>% 
  group_by(year,throws) %>% 
  count() %>%
  ungroup() %>% 
  drop_na() %>% 
  group_by(year) %>% 
  mutate(
    "N" = sum(n),
    "proportion" = round(n/sum(n),3)
  ) 

```

##Plot 
Again, ```ggplotly()``` was used to add the number of total observations a given years proportional estimate is based upon as a text element.   

```{r}
throwPlot <- 
  ggplot(plrThrows, aes(year, proportion, label = N))+
  geom_line(aes(color=throws))+
  scale_x_continuous(name="Year", breaks = seq(1871,2018,20))+
  scale_y_continuous(breaks=seq(0,1,0.1))+
  scale_color_brewer(palette="Set1")+
  ylab("Proportion of Preferred Throwing Hand")+
  labs(color="Throwing \n Hand")+
  ggtitle("Proportional Distribution of Throwing Hands")

ggplotly(throwPlot,tooltip = c("proportion", "year", "N"))
```    
Again, the plot might have us think that the proportion of right handed throwers decreased over time (while the proportion of left handed throwers increased), yet, a more thorough inspection of the single data points reveals that the early proportional estimates are based on relatively small sample sizes.

##Wins Above Replacement and Hand Preference in 2016     

Wins Above Replacement (WAR) is an often used "sabermetric" to quantify how valuable a given player is to his team's success. It is the number of a team's additional wins that are due to the player as compared to the number of expected wins if the player had been replaced by a hypothetical average player. The computation of WAR is quite complex, and, depending on whether we're dealing with a position player or a pitcher, WAR is computed differently. Luckily, baseball-reference.com provides WAR statistics for (alledgedly) every player that has ever played in the league taking into account a player's position (i.e., there are two data tables: One for position players and one for pitchers; Link: https://www.baseball-reference.com/about/war_explained.shtml).  
We downloaded Baseball Reference's "War for Position Players" (```warPos```) and "War for Pitchers" (```warPit```) tables and combined these with Lahman's baseball data bank to use WAR as the outcome variable of our linear models (see below).   

All of our linear models were restricted in that we only used data from the 2016 season. 

###Handedness and WAR for Position Players  
The aim here is to compare position players' WAR statistics as a function of batting and throwing hand. 

First, we used ```filter()``` to restrict our WAR and Appearances data tables to observations in the 2016 season. 
```{r}
#Select subset of position players that played in 2016
warPos2016 <- warPos %>% 
  filter(year_ID==2016)

#Select subset of players that played in 2016
appearances2016 <- Appearances %>% 
  filter(yearID == 2016)  
```

Next, we identified players who played for more than one team in 2016, the reason being is that these players would show up as multiple observations (i.e., multiple rows) in the ```warPos2016``` data table. Also, we checked whether this conjecture of individual players showing up more than once in the table is in fact due to them playing for more than one team. 
```{r}
#Show players that played for multiple teams in 2016
warPos2016 %>% 
  count(player_ID) %>% 
  arrange(desc(n)) %>% 
  filter(n>1)

#Check if player_ID-by-teamID is a unique key in warPos2016
warPos2016 %>% 
  count(player_ID, team_ID) %>% 
  filter(n>1)
```  

In order to merge ```warPos2016``` and ```appearances2016``` we again needed a unique key. Unfortunately, the column "playerID" in the ```appearances2016``` table does not match the column "player_ID" in the ```warPos2016``` data table. There is, however, a column named "bbrefID" in the ```People```which corresponds to the "player_ID" column in the the ```warPos``` and ```warPit``` tables. Note that the ```players``` table created earlier already contains the "bbrefID" column. Thus, we merged ```appearances2016``` with ```players``` in order to add the "bbrefID" column.  

```{r}
#Get the bbrefID column
pos2016 <- appearances2016 %>% 
  left_join(players, by="playerID")
```

Again, we checked whether the combination of "bbrefID" and "teamID" served as a unique key.
```{r}
#Check if bbrefID-by-teamID is a unique key in appearances2016
pos2016 %>% 
  count(bbrefID,teamID) %>% 
  filter(n>1)
```

We then joined ```appearances2016``` and ```warPos2016``` using player and team IDs as keys. 
```{r}
#Combine appearances2016 and warPos2016 tables
pos2016 <- pos2016 %>% 
  left_join(warPos2016, by = 
              c(c("bbrefID" = "player_ID"), c("teamID"="team_ID"))) %>% 
  select(playerID, teamID, G_all, WAR, bats, throws)
```

For our linear models, we needed each player to be represented by exactly one row in the ```appearances2016``` table (i.e., exactly one WAR per player). Since some players played for more than one team, and, for each team they had (slightly) different WARs we decided to weigh the player's WARs by the proportion of games he played for each team and take the sum (i.e., a weighted mean WAR):
$$
\text{weightedWAR}_i = \sum_{k=1}^{m}\frac{\text{number of games played for team }k }{\text{total number of games played}}\times WAR_k ,
$$  
where $i$ refers to the $i$th player, and $k$ to the teams he played for in 2016. Note that if a player's WAR for a given team was coded as ```NA```, the player's stint with that team was simply ignored in the computation. This step is reflected in the ```drop_na()``` command in the function ```wghtWAR()```. 


In order to apply the ```wghtWAR()``` function to each player, we first nested WAR within players. We then used ```lapply()``` to compute the weighedWAR statistic for each player. 
```{r}
#Nest number of WAR-osbervations by players
pos2016 <- nest(pos2016, c(G_all,teamID), key = "WAR")

#Compute weighted WAR for each player
pos2016 <- pos2016 %>% 
  mutate(
  "weightedWAR" = lapply(pos2016$data, function(x) wghtWAR(x)) %>% unlist
)   
```

We then plotted "weightedWAR" as a function of batting hand.
```{r}
#raw data for batting hand
pos2016 %>% 
  drop_na() %>% 
  ggplot()+
  geom_jitter(aes(bats,weightedWAR, color=bats),alpha=.5,
              position=position_jitter(width=.2))+
  scale_color_brewer(palette="Set1")
```

```{r}
#raw data for throwing hand
pos2016 %>% 
  drop_na() %>% 
  ggplot()+
  geom_jitter(aes(throws,weightedWAR, color=throws),alpha=.5,
              position=position_jitter(width=.2)) +
  scale_color_brewer(palette="Set1")
```


###WAR for Pitchers
```{r}
#Select subset of pitchers that played in 2016
warPit2016 <- warPit %>% 
  filter(year_ID==2016)

#Add bbrefID column to pit2016
pit2016 <- appearances2016 %>% 
  left_join(players, by="playerID")

#Combine appearances2016 and warPit2016 tables
pit2016 <- pit2016 %>% 
  left_join(warPit2016, by = 
              c(c("bbrefID" = "player_ID"), c("teamID"="team_ID"))) %>% 
  select(playerID, teamID, G_all, WAR, bats, throws)

#Nest number of WAR-osbervations by players
pit2016 <- nest(pit2016, c(G_all,teamID), key = "WAR")

#Compute weighted WAR for each player
pit2016 <- pit2016 %>% 
  mutate(
  "weightedWAR" = lapply(pit2016$data, function(x) wghtWAR(x)) %>% unlist
)   
#raw data for batting hand
pit2016 %>%   
  drop_na() %>% 
  ggplot()+
  geom_jitter(aes(bats,weightedWAR, color=bats),alpha=.5,
              position=position_jitter(width=.2))+
  scale_color_brewer(palette="Set1")

#raw data for throwing hand
pit2016 %>%   
  drop_na() %>% 
  ggplot()+
  geom_jitter(aes(throws,weightedWAR, color=throws),alpha=.5,
              position=position_jitter(width=.2)) +
  scale_color_brewer(palette="Set1")
```


# Of Age and Baseball

Let's anwers the simplest question first.

## How old is the average active professional baseball player?

```{r}
age_calc <- function(from,to=now()) floor(difftime(to, from)/dyears(1))
age_year = 2016
age_const = paste0(age_year,"0403")

age_c_players <- 
    left_join(Appearances %>% filter(yearID == age_year), People, "playerID") %>%
    mutate(birthDate = as_date(ymd(paste(birthYear, birthMonth, birthDay, sep = "-")))) %>%
    mutate(age = age_calc(birthDate, ymd(age_const))) %>%
    select(playerID, teamID, birthDate, age, nameFirst, nameLast, G_all, GS) 
```

We define active as in "played this season" which in our case and dataset is the season of `r age_year`. Also we calculate the age with the start of the season (April 3rd).

We find that the average active professional baseball player , in the season of `r age_year`, was born `r mean(age_c_players$birthDate)` and was therefore `r age_calc(mean(age_c_players$birthDate),ymd(age_const))` years old. 

Let's visualize that.

```{r}
age_c_players %>% 
    ggplot(aes(x = age)) +
    geom_line(stat = "density") + 
    geom_vline(aes(xintercept = mean(age)), alpha = 0.5, color = "red")
```

Funfact: As we might be able to tell from the graph, the oldest player in the season of `r age_year` was `r age_calc(min(age_c_players$birthDate),ymd(age_const))`.

Now we know about the average active player. Let's go further.

## Does the age distribution differ significantly per team?

```{r}
names(color$team_main) <- sort(unique(age_c_players$teamID))

age_c_team <- age_c_players %>% 
   group_by(teamID) %>%
   summarise(meanAge = mean(age), medianAge = median(age)) %>%
   left_join(Teams %>% filter(yearID == age_year) %>% select(teamID,name,W,L,G), by = "teamID")
 
age_c_team %>% 
    ggplot(aes(reorder(teamID,meanAge), meanAge, color = teamID)) +
    geom_point(size = 2) + 
    scale_color_manual(values = color$team_main) +
    guides(color = FALSE) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
```
So now we know that the the team with the - on average - youngest players are the `r (age_c_team %>% filter(meanAge == min(meanAge)))$name` while the `r (age_c_team %>% filter(meanAge == max(meanAge)))$name` deploys -again, on average- the oldest players. But the difference between these 'extremes' is `r signif(max(age_c_team$meanAge) - min(age_c_team$meanAge),3)` and in my view not that siginificant. 

We still want to know more, let's look at the related boxplot.

```{r}
age_c_players %>% 
    left_join(age_c_team, by = "teamID") %>%
    ggplot(aes(reorder(teamID,meanAge),age,color = teamID)) +
    geom_boxplot(notch = TRUE) +
    scale_color_manual(values = color$team_main) +
    guides(color=FALSE) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
```
As we see, the IQR and Median isn't _that_ different between the teams. Maybe we can work out larger differences in the next question.

## What happens to the average team age if we weight the games that each player played?

The idea is to weight the playing time of each player and therefore see if teams rather put their faith in younger players or rather value experience. The formula we use to weigh each time is $$ \text{age}_p \frac{\text{games played}_p}{\frac{1}{n}\sum^n_{i=1}\text{games played}_i} $$

```{r}
age_c_team <- 
   left_join(
             age_c_players,
             (Teams %>% filter(yearID == age_year) %>% select(teamID,G, W, L)),
             "teamID"
            ) %>%
   group_by(teamID) %>% 
   add_count() %>%
   mutate(G_avg = mean(G_all),GS_avg = mean(GS)) %>%
   mutate(gProp = age*(G_all/G_avg), gsProp = age*(GS/GS_avg)) %>%
   summarise(weightByG = mean(gProp), weightByGS = mean(gsProp)) %>%
   left_join(age_c_team, "teamID")

#kable(sample_n(age_c_team,5), format="markdown")

age_c_team %>% 
    ggplot(aes(reorder(teamID, meanAge), weightByG, color = teamID)) +
    geom_point(aes(reorder(teamID, meanAge), meanAge, color = teamID), size = 0.8) +
    geom_point(size = 2) + 
    scale_color_manual(values = color$team_main) +
    geom_segment(aes(xend = teamID, yend = meanAge)) +
    guides(color = FALSE) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
```

As we can can see, there are some differences if we weight the playing time, but most noticeable to me is that the shift is clearly toward more experienced players. To be exact, according to this new metric the average team is now `r signif(age_c_team %>% summarise(mean(weightByG-meanAge)),4)` years older which adds up to `r signif(age_c_team %>% summarise(sum(weightByG-meanAge)),4)` years in total for the whole league.

## Linear Model - performance and average team age

```{r}
age_c_team <- age_c_team %>% mutate(Wprop = W/G)
age_perf_mod <- lm(data=age_c_team, formula= Wprop ~ weightByG)
age_c_team <- age_c_team %>% add_predictions(age_perf_mod) %>% add_residuals(age_perf_mod)

age_c_team %>% ggplot(aes(weightByG)) + geom_point(aes(y=Wprop)) + geom_line(aes(y=pred),color="red")

age_c_team %>% ggplot(aes(resid)) + geom_freqpoly(binwidth=0.04)

age_c_team %>% ggplot(aes(weightByG, resid)) + geom_point()
```

```{r results="asis"}
stargazer(age_perf_mod, align=T, type='html')
```

## performance of individuals as they age

Adjusting the US-Salary statistics
The statistics for the average wage in the US came from the OECD (https://data.oecd.org/earnwage/average-wages.htm#indicator-chart). Unfortunately, we didn't find more accurate data for the wages than the average. Be careful with interpreting the numbers, as with wages, the mean and the median differ a lot. 
Within the dataset some columns were dropped, and the average development in percent was calculated. 
```{r}
#renaming the columns
colnames(Salaries_US) = c("country", "indicator", "subject", "currency", "frequency", "yearID", "av_wage_US", "Flag_codes")
Salaries_US <- Salaries_US %>%
  #dropping columns
  subset(select = -c(country, indicator, subject, frequency, Flag_codes)) %>%
  #calculating average development
  mutate(av_wage_perc = ((av_wage_US/lag(av_wage_US) - 1) * 100),
         av_wage_100perc = ((av_wage_US-min(av_wage_US))/max(av_wage_US)))
(Salaries_US)
```
  
# Salaries
Adjusting the Baseball-Salary statistics
With the baseball salaries, some statistical values, like mean, max, min and median were calculated. Also the average development of the mean salary was computed to compare it to the US salaries. 
```{r}
sal_bb_stats <- Salaries %>%
  group_by(yearID) %>%
  summarise(n_players = n(), 
         n_dist_players = n_distinct(playerID),
         salary_mean = mean(salary),
         salary_max = max(salary),
         salary_min = min(salary),
         salary_median = quantile(salary, probs = 0.5))

#
sal_bb_stats_90 <- sal_bb_stats %>%
  filter(yearID >= 1990) %>%
  mutate(salary_mean_perc = ((salary_mean/lag(salary_mean) - 1) * 100),
         salary_mean_100perc = ((salary_mean-min(salary_mean))/max(salary_mean))) 
```
  
##Salaries overview
The boxplot shows the median salary, and the first and third quantile. There are a lot of outliers. The regression line is around the mean salary, which is also shown. In 2000 there was a leap in the median salary, and in and after 2009, the highest salaries so far were payed. We're going to take a look on that afterwards. 

```{r}
summary(Salaries)

#boxplot with the player salaries
sal_boxplot <- ggplot() + 
  geom_boxplot(data = Salaries, aes(x = yearID, y = salary, group = cut_width(yearID, 1)), alpha = 0.5) + 
  geom_smooth(data = Salaries, aes(x = yearID, y = salary), method='lm', se=F) +
  geom_line(data = sal_bb_stats, aes(x = yearID, y = salary_mean)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = Salaries$yearID) +
  labs(x = "Year", y = "Salary") 
(sal_boxplot)

#graph with number of players per year
sal_bb_nbplayers <- ggplot(data = sal_bb_stats, mapping = aes(x = yearID, y = n_players)) + 
  geom_line() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = Salaries$yearID) +
  labs(x = "Year", y = "Number of players")
(sal_bb_nbplayers)
```

##Salaries within the leagues
Let's focus on the differences within the leagues. We have 2 leagues within the salaries data, the National League (NL) and the American League (AL). In the scatterplot we see, that the differences between the average salary and maximum salary grow with time, so this is not caused by any league paying highly more, although there are differences: The maximum gaining players of the NL earn more money than the ones from the AL (except for 2015 and 2016). 
But over the average salaries, the AL pays more. 
So if you're good, go the the AL - if you're the best, go to the NL. 
The range, which tells us about the gap between the min and max salary of the players rose over time, and here we can see, that the least paid player gets a lot less compared to the maximum paid player in the AL than in the NL. 
```{r}
#computing the stats for leagues 
sal_bb_stats_lg <- Salaries %>%
  group_by(yearID, lgID) %>%
  summarise(n_players_lg = n(), 
         salary_lg_mean = mean(salary),
         salary_lg_max = max(salary),
         salary_lg_min = min(salary),
         salary_lg_median = quantile(salary, probs = 0.5)) %>%
  mutate(salary_lg_100perc = (salary_lg_mean-min(salary_lg_mean)/max(salary_lg_mean)))
(sal_bb_stats_lg)

#graph with the salaries, grouped by league
sal_lg <- 
  ggplot(data = Salaries, mapping = aes(x = yearID, y = salary)) + 
  geom_point(mapping = aes(color = lgID), alpha = "0.3", position = "jitter") + 
  geom_smooth(method='lm', se=F) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y = "Salary")
(sal_lg)

#graph with the salaries, grouped by league: boxplot
#sal_lg_boxplot <- 
  #ggplot(data = sal_bb_stats_lg, mapping = aes(x = yearID, y = salary)) + 
  #geom_boxplot(aes(colour = lgID, group = cut_width(yearID, 1)), alpha = 0.5) + 
  #theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  #scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  #labs(x = "Year", y = "Salary")

#graph with number of players per year and league
#ggplot(data = sal_bb_stats_lg, mapping = aes(x = yearID, y = n_players_lg)) + 
  #geom_line(mapping = aes(color = lgID))+ 
  #theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  #scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  #labs(x = "Year", y = "Number of players")

#graph with average salaries per year and league
sal_bb_stats_lg_mean <- 
  ggplot(data = sal_bb_stats_lg, mapping = aes(x = yearID, y = salary_lg_mean)) + 
  geom_line(mapping = aes(color = lgID)) +
  geom_point(mapping = aes(color = lgID)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y = "Average salary (mean)")
(sal_bb_stats_lg_mean)

#graph with max, min and median and mean salary per year and league
sal_bb_stats_lg_all <- 
  ggplot(data = sal_bb_stats_lg, mapping = aes(shape = lgID, color = lgID), size = 10) +
  geom_line(mapping = aes(x = yearID, y = salary_lg_max)) +
  geom_line(mapping = aes(x = yearID, y = salary_lg_min)) + 
  geom_line(mapping = aes(x = yearID, y = salary_lg_median)) +
  geom_line(mapping = aes(x = yearID, y = salary_lg_mean)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y ="Salary (max, min, median, mean)")
(sal_bb_stats_lg_all)

#Range of bb salaries in leagues
sal_bb_stats_lg_range <-
  ggplot(data = sal_bb_stats_lg, mapping = aes(color = lgID)) +
  geom_line(mapping = aes(x = yearID, y = salary_lg_max-salary_lg_min)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y = "Range of salaries in US$", 
       title = "Range of salaries of baseball players by league")
(sal_bb_stats_lg_range)
```

##Average baseball and US salaries
The average baseball salaries increased a lot with time. Let's compare it to the average US salaries and see if they rose alike. 
We can see, that the average salary of baseball players is a lot higher. To see the differences more, we compare the percentage rise over time. Starting at 1990 (because the US Salary data is not available for the years before), the average US wages increased about 25% until 2016, whereas the baseball players' salaries increased more than 85%. This is a huge difference. So, baseball probably became more popular or at least, people want to spend more on the games and the teams, the teams might have more money because of bigger sponsoring and so on. 

```{r}
sal_bb_us <- left_join(sal_bb_stats_90, Salaries_US, by = "yearID")
(sal_bb_us) 

#graph with average salaries per year for bb
sal_bb_us_av <- 
  ggplot(data = sal_bb_us) + 
  geom_point(mapping = aes(x = yearID, y = salary_mean), color = "forestgreen") + 
  geom_smooth(mapping = aes(x = yearID, y = salary_mean), method='lm', se=F) + 
  geom_point(mapping = aes(x = yearID, y = av_wage_US), color = "deeppink2") + 
  geom_smooth(mapping = aes(x = yearID, y = av_wage_US), method='lm', se=F) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_us$yearID) +
  labs(x = "Year", y = "Average salary", title = "Average salary of baseballers and US citizens")
(sal_bb_us_av)

sal_bb_us_dev <- ggplot(data = sal_bb_us) + 
  geom_point(mapping = aes(x = yearID, y = salary_mean_100perc), color = "forestgreen") + 
  geom_smooth(mapping = aes(x = yearID, y = salary_mean_100perc), method='lm', se=F) + 
  geom_point(mapping = aes(x = yearID, y = av_wage_100perc), color = "deeppink2") + 
  geom_smooth(mapping = aes(x = yearID, y = av_wage_100perc), method='lm', se=F) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_us$yearID) +
  labs(y = "Percentage rise in salary of baseballers and US citizens", title = "Salarydevelopment of baseballers and US citizens")
(sal_bb_us_dev)
```
  
##Mostpaid players and mostpaying teams

The rise in the salaries was mainly conducted by some high earninig players - why do they get paid so much more than others? Are they better? To answer these questions the players with the largest salaries each year (after 2000, because after that year, the gap was getting a lot bigger) were chosen for further analysis. 
```{r}
salary_max <- Salaries %>%
  filter(yearID > 1999) %>%
  group_by(yearID) %>%
  mutate(max_salary = max(salary)) %>%
  subset(salary == max_salary) %>%
  ungroup()
(salary_max)

mostpaid_players <- salary_max %>%
  group_by(playerID) %>%
  summarise(mostpaid_player = n()) %>%
  ungroup()
(mostpaid_players)

mostpaying_teams <- salary_max %>%
  group_by(teamID) %>%
  summarise(mostpaying_team = n())
(mostpaying_teams)

sal_mostp_pl <- Salaries %>%
  filter(playerID %in% c(salary_max$playerID))
(sal_mostp_pl)

sal_mostp_team <- Salaries %>%
  filter(teamID %in% mostpaying_teams$teamID)
(sal_mostp_team)

```

###Mostpaid players
```{r}
sal_mostp_pl_all <- 
  ggplot(data = sal_mostp_pl, mapping = aes(x = yearID, y = salary)) + 
  geom_line(mapping = aes(color = playerID)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y = "Salary per player in US$")
(sal_mostp_pl_all)  

```

###Mostpaid teams
```{r}
sal_mostp_team_stat <- sal_mostp_team %>%
  group_by(yearID, teamID) %>%
  summarise(n_players = n(), 
         salary_mean = mean(salary),
         salary_max = max(salary),
         salary_min = min(salary),
         salary_median = quantile(salary, probs = 0.5)) %>%
  mutate(salary_100perc = (salary_mean-min(salary_mean)/max(salary_mean)))
(sal_mostp_team_stat)

sal_mostp_teams_all <- 
  ggplot(data = sal_mostp_team_stat) + 
  geom_line(mapping = aes(x = yearID, y = salary_mean, color = teamID)) + 
  geom_line(mapping = aes(x = yearID, y = salary_median, color = teamID)) + 
  geom_line(mapping = aes(x = yearID, y = salary_max, color = teamID)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  scale_x_continuous(breaks = sal_bb_stats_lg$yearID) +
  labs(x = "Year", y = "Mean, max, median salary per team")
(sal_mostp_teams_all)  

```
